Theoretical prediction of disrupted Min oscillation in flattened \emph{Escherichia coli}

### 2015-08-01 paper/response-to-referees.txt

Referee #1.

1. The largest concern of Referee #1 was our choice of Huang et al's
model for our simulations.  The referee acknowledges that we can stay
with this model, but feels that our justifaction that this model is
inadequate.  The referee objects that the fact that this model has
been widely is an inappropriate justification.  While we agree that
this alone is not a sufficient reason for studying this model, we do
believe that it is relevant: this is not some obscure older model that
we are testing, but rather the basis of all more recent models.  We
believe that the limitations of this model are of interest, and have
added some discussion to the end of the introduction section
explaining this.

2. As a secondary point, the referee points out we cannot conclude
that some other deterministic model could not demonstrate chaotic
behavior that explains the experimental results of Mannik et al.  We
agree that this cannot be ruled out, and have added mention of this to
our "Comparison of Experimental and Stadium Shapes", and have made
even more tentative our suggestion in the summary that other similar
deterministic models may also prove to be non-chaotic.

3. The referee when looking at Figure 4 observes that the stadium
shapes exhibit oscillation that is largely bipolar, and expresses
doubt in our claim that the oscillations in stadium shapes are
irregular.  In particular, the referee points out that if we average
over many periods, we would find that the oscillations are indeed
bipolar.  We agree with this statement, but not with the conclusion.
Averaging the locations of the maxima would indeed miss the point of
the comparison with experiment, which is in examining the sequence of
locations of maxima.

@@

5. Based on the referee's suggestion, we have clarified that reference
11 does not extend Huang's model, except in that it simulates the same
reaction-diffusion equations stochastically.

6. The referee points out that we failed to mention that stochastic
simulations are needed to explain the experimentally observed
length-dependence of fluctuations in the Min system.  We have added
mention of this additional evidence of the importance of inclusion of
stochastic effects, in the paragraph suggested by the referee.

Referee #3.

7. The referee #3 also expressed doubt about our conclusion with
regard to the effect of irregular shape which we base on our stadium
models.  Unlike the previous referee, the criticism is based not on
the spatial output (Figure 4), but rather on the reported coherence
times.  We agree with the referee that our concluding sentence
immediately prior to the conclusion section was unjustified, and have
edited this sentence to reflect the referee's objections.

8. The referee also questioned how the cell was divided in order to
perform the coherence time calculations, pointing out that a smaller
region could lead to a shorter coherence time.  We have added an
explanation of how we divided the cell into three sections, and gave
numbers showing the relative volume of the sections.  These were not
identical, and thus will have some effect on the correlation times
computed.  However, the region volumes are roughly similar, so we do
not expect this effect to be dramatic.

9. The referee points out that the "starry night" patterns are
inconsistent with experiments.  We agree with the referee that
membrane diffusion will alleviate this problem.  We have added a
clarification that this effect is *not* observed in experiment.

@@

11. The referee suggests that we should consistently write $\mu m^2$
over $\mu^2$ for square microns.  We have made this change to the plot
that used $\mu^2$.

12. The referee suggested a title is needed in the caption of Figure
number 4.  We have edited the caption to begin with an appropriate
title-like sentence fragment, rather than "We display here...".

13. As requested we have corrected the capitalization of "E. coli".

14. The referee corrected us regarding the location at which the FtsZ
polymer forms, which is not on the cell wall, but rather near the
cytoplasmic side of the inner membrane.  We have corrected this
statement in the first paragraph of the introduction.

### 2015-08-01 paper/plos.tex
edits while meeting with Jeff and Rene

* E-mail: Corresponding schuljef@physics.oregonstate.edu
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
  The dynamics of the Min-protein system help \emph{Escherichia coli}
  regulate the process of cell division by identifying the center of
  the cell.  While this system exhibits robust bipolar oscillations in
  wild-type cell shapes, recent experiments have shown that when the
  cells are mechanically deformed into wide, flattened out, irregular
  shapes, the spatial regularity of these oscillations breaks down. We
  employ widely used stochastic and deterministic models of the Min
  system to simulate cells with flattened shapes.  The deterministic
  model predicts strong bipolar oscillations, in contradiction with
  the experimentally observed behavior, while the stochastic model,
  which is based on the same reaction-diffusion equations, does
  predict spatially irregular oscillations\fixme{, although they are still somewhat more bipolar than seen in experiment}.
  We further report simulations of flattened but more symmetric
  shapes, and find that it \fixme{may be} the flattening and lateral expansion
  rather than the asymmetry of the cell shapes that causes the
  irregular oscillation behavior.

@@

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
the flattening and accompanying lateral expansion of the cell versus
cell shape irregularity and asymmetry in the temporal regularity of
the oscillations.  Although in the movies the stadiums appear to
display, on average over long time scales, a somewhat more regular
oscillation than appears in the irregular shapes \emph{A} and
\emph{B}, the limited time range of the arrow plots (350 seconds) is
inadequate to make this comparison.  We therefore return to the
correlation function between the number of proteins in a segment at
the top of the cell and the number of proteins in a segment at the
bottom of the cell.  \fixme{this is where Ill need to add the volumes} Fig.~\ref{fig:corr-B} shows this plot for shape
\emph{B} and stadium \emph{B}.  Here we plot correlations of
correlation times up to 500 seconds, a biologically relevant
timescale.  E. coli cell doubling times vary widely depending on
environmental conditions, but will often be between 20 minutes to 100
minutes~\cite{pierucci1972chromosome}, while fluorescent microscopy
experiments have shown that the FtsZ ring builds itself with a half
life of roughly 30 seconds~\cite{stricker2002rapid}.

@@

\section*{Conclusion}
\fixme{We find a certain amount of qualitive agreement between our stochastic simulations and experiments showing disrupted MinD oscillations in asymmetric flattened E. coli cells.}  We find agreement between our stochastic simulations and experiments
showing disrupted bipolar MinD oscillation in asymmetric flattened
E. coli cells~\cite{mannik2012robustness}.  As observed
experimentally, our simulations predict that MinD maxima will form in
a spatially irregular sequence of locations, \fixme{albeit with a somewhat more course scale bipolar regularity.}  This result builds upon
existing results showing that this model and its variants are
effective in a variety of wild-type and mutated cell
shapes~\cite{fange2006noise, varma2008min, kruse2007experimentalist},
and reinforces that stochastic variations of Huang's 2003
model~\cite{fange2006noise, kerr2006division} has considerable
predictive power beyond the standard wild-type cell.

>>>>

* E-mail: Corresponding schuljef@physics.oregonstate.edu
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
  The dynamics of the Min-protein system help \emph{Escherichia coli}
  regulate the process of cell division by identifying the center of
  the cell.  While this system exhibits robust bipolar oscillations in
  wild-type cell shapes, recent experiments have shown that when the
  cells are mechanically deformed into wide, flattened out, irregular
  shapes, the spatial regularity of these oscillations breaks down. We
  employ widely used stochastic and deterministic models of the Min
  system to simulate cells with flattened shapes.  The deterministic
  model predicts strong bipolar oscillations, in contradiction with
  the experimentally observed behavior, while the stochastic model,
  which is based on the same reaction-diffusion equations,
  predicts more spatially irregular oscillations.
  We further report simulations of flattened but more symmetric
  shapes, which suggest that the flattening and lateral expansion may
  contribute as much to the irregular oscillation behavior as the
  asymmetry of the cell shapes.

@@

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
the flattening and accompanying lateral expansion of the cell versus
cell shape irregularity and asymmetry in the temporal regularity of
the oscillations.  Although in the movies the stadiums appear to
display, on average over long time scales, a somewhat more regular
oscillation than appears in the irregular shapes \emph{A} and
\emph{B}, the limited time range of the arrow plots (350 seconds) is
inadequate to make this comparison.  We therefore return to the
correlation function between the number of proteins in a segment at
the top of the cell and the number of proteins in a segment at the
bottom of the cell.  The ``top'' and ``bottom'' regions for computing
the correlation time were established by dividing the long axis into
thirds.  This resulted in regions at the top and bottom ranging from
19\% to 33\% of the cell volume for shapes \emph{A} and \emph{B}, with
the stadium shapes featuring 30\% of the volume in the top and bottom
regions.  Fig.~\ref{fig:corr-B} shows this plot for shape \emph{B} and
stadium \emph{B}.  Here we plot correlations of correlation times up
to 500 seconds, a biologically relevant timescale.  E. coli cell
doubling times vary widely depending on environmental conditions, but
will often be between 20 minutes to 100
minutes~\cite{pierucci1972chromosome}, while fluorescent microscopy
experiments have shown that the FtsZ ring builds itself with a half
life of roughly 30 seconds~\cite{stricker2002rapid}.

@@

\section*{Conclusion}
We find qualitative agreement between our stochastic simulations and
experiments showing disrupted bipolar MinD oscillation in asymmetric
flattened E. coli cells~\cite{mannik2012robustness}.  As observed
experimentally, our simulations predict that MinD maxima will form in
a spatially irregular sequence of locations, which follow an overall
bipolar trend.  This result builds upon existing results showing that
this model and its variants are effective in a variety of wild-type
and mutated cell shapes~\cite{fange2006noise, varma2008min,
  kruse2007experimentalist}, and reinforces that stochastic variations
of Huang's 2003 model~\cite{fange2006noise, kerr2006division} has
considerable predictive power beyond the standard wild-type cell.
